Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-05 tagless/tagless-final.tex
more of the related work

Our initial motivation came from several papers 
\citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
embedded interpreters as a central example to justify looking 
into advanced type systems.  While we admired all this technical work,
there remained the issue that we were not convinced of the strength
of their motivating example.

\citet{WalidICFP02}, when referring to the work of \citet{taha-tag},
claims that this shows that self-interpretation is possible in a
\emph{simply typed $\lambda$-calculus}, but this is somewhat misleading
as their type system does not differentiate between the various ground
types.  In other words, various constants (like \textsf{nil} and 
\textsf{true}) both inhabit the same type $D$.  This is not very satisfying.

\oleg{He did not eliminate all the tags.}\jacques{can you show me?}
\oleg{WalidICFP02, the very end of p1 and the beginning of p2, first
  column}
From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
impression is that the claim is:
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type,
\item It is not possible to write a \emph{direct} interpreter for an typed
language encoded this way, in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus we have to use an indirect interpreter, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the first step
was not valid.

From the above:
For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''

@@

Asai; 
It [dynamic part when dealing with bool, int, lam in PE] is
Asai's trick for reducing the amount of reify/reflect back-and-forth
in a PE for an untyped language that allows you back-and-forth.  Or at
least that's how I read those papers.
Asai's approach is related to our PE (emulating typelevel functions):
Ken: ``In some sense we're just observing that Asai's code type-checks
-- in Hindley-Milner as soon as we deforest the static code
representation.''

Sumii et al.

"Boxes go Bananas" by Washburn and Weirich (intended or otherwise) Ken
wrote about this: ``but a glance reveals no type classes.  Do you mean
then that if we hide the definition |newtype R a = R a| from a client,
and export only the interpreter |instance Symantics R R|, then the
client would see R as an HOAS datatype a la Washburn and Weirich and
only be able to create parametric ("non-exotic") terms?''  Jacques
commented: ``I mean the HOAS parts, and how to deal with it (like in
the instance for |Symantics R R|).  There are other nice tricks for
dealing with HOAS in that paper worth a closer look.  It's all about
plain polymorphism, not type classes, but there are still some clever
ideas there, and they are about issues of tagging (an lack thereof).''

Typing Dep Typing plus two messages on Haskell-Cafe: typechecking 
from untyped to the typed form, ready to be interpreted. Walid's ICFP02
uses dependent types
\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019160.html}
\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019161.html}
\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019168.html}

Regarding our implementation of the type-level function |'sv -> 'dv|.
Gibbons et al paper on
the implementations of typecase may be relevant. I have not read it,
but I have the feeling some of their typecase designs may be
relevant (at least to cite). 
However, the typecase paper relies on having a 
bijection between types -- our |'sv| and |'dv| in this case, which is not 
our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
that |'sv|.  \jacques{Could we say that we have a meta-level Galois
injection?  We have one direction completely covered at the object
level (and so easy to lift), but 'dv to 'sv can only be done by
an outside mechanism}

@@

Thiemann, cogen in six lines (for the interpreter/compiler work)?

>>>>

Our initial motivation came from several papers 
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
embedded interpreters as a central example to justify looking 
into advanced type systems.  While we admire all this technical work,
there remained the issue that we were not convinced that their 
motivating example actually needed all this machinery.

\citet{WalidICFP02}, when referring to the work of \citet{taha-tag},
claims that this shows that self-interpretation is possible in a
\emph{simply typed $\lambda$-calculus}, but this is somewhat misleading
as their type system does not differentiate between the various ground
types.  In other words, various constants (like \textsf{nil} and 
\textsf{true}) both inhabit the same type $D$.  This is not very satisfying.
Furthermore, as is pointed out in the introduction to \citet{WalidICFP02},
\citet{taha-tag} does not \emph{statically} guarantee that all tags will
be removed - tag elimination must be performed at runtime.

From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
impression is that the claim is:
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the very first step
was not valid.

In particular, \citet[\S5]{taha-tag}
states that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''

@@

We used a method from \citet{Asai-SAS99} to reduce the amount
of back-and-forth between reify/reflect: in our PE we track both
static and dynamic values in parallel (whenever possible).  In some sense,
we are observing that this method type-checks in Hindley-Milner once we
deforest the static code representation.  Put another way, we have
manually applied type-level partial-evaluation to our typelevel 
functions (see \S\ref{S:pe-solution}) to obtain simpler types 
acceptable to MetaOCaml.  It is worthwhile to note that 
\citet{Sumii-HOSC2001} also takes advanatage of this method
\citep[of][]{Asai-SAS99} to get interesting results combining online
and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
we strive for modularity by reusing earlier stages.  It would be interesting
to see if we could derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.

The approach to HOAS via catamorphism and anamorphism of
\citet{Washburn-Weirich-boxes} seems related to what we are doing.
We have not investigated the relationship to any serious depth, but we
feel that one could fruitfully combine their ideas with ours.

A lot of effort has gone into finding ways to give precise types to untyped
terms (see for example \citet{baars-typing,Guillemette-Monier-PLPV}
as well as the messages \citet{haskell-019160,haskell-019161}).  There 
take advantage of the host language's types to varying extents.  All 
these approaches are very successful for proper terms, but less so for
ill-typed terms, where error messages can be quite obscure.  We use the
host's typing facilities to do all the work, and thus inherit the strengths
and weaknesses of the host language.  While that may not be optimal, it is
certainly much more modular.

\jacques{Got to here}.
Regarding our implementation of the type-level function |'sv -> 'dv|.
Gibbons et al paper on
the implementations of typecase may be relevant. I have not read it,
but I have the feeling some of their typecase designs may be
relevant (at least to cite). 
However, the typecase paper relies on having a 
bijection between types -- our |'sv| and |'dv| in this case, which is not 
our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
that |'sv|.  \jacques{Could we say that we have a meta-level Galois
injection?  We have one direction completely covered at the object
level (and so easy to lift), but 'dv to 'sv can only be done by
an outside mechanism}

@@

co-algebra references.  Many.
